Convergence Studies and Effective Alphas for Disks with Gravitational Instabilities. \\ I. The Constant Cooling Case

### 2010-12-08 Paper_1.tex
Rearranged figure labels.
	modified:   Paper_1.tex
	renamed:    figures/qcompareres.eps -> figures/Figure1.eps
	new file:   figures/Figure2.eps
	new file:   figures/Figure3.eps
	new file:   figures/Figure4.eps
	new file:   figures/Figure5.eps
	new file:   figures/Figure6.eps
	new file:   figures/Figure7.eps
Tom's edits
	modified:   Paper_1.tex

\title{Convergence Studies and Effective Alphas for Disks with Gravitational Instabilities.
 \\
I. The Constant Cooling Case}
\author{Scott Michael}
\affil{Astronomy Department, Indiana University, Bloomington, IN 47405}
\email{scamicha@astro.indiana.edu}

@@

Using 3D hydrodynamic simulations of a cooling protostellar disk undergoing gravitational instabilities (GIs), we compute the effective \citet{shakura1973} $\alpha$ due to gravitational stresses and compare them to predictions from an analytic local theory for thin disks by \citet{gammie2001}.  Our goal is to determine how accurately a locally defined $\alpha$ can characterize mass and angular momentum transport by GIs in disks.  We examine disks with an imposed constant global cooling time \citep{mejia2005}.  Models were computed with grids of different spatial resolution to investigate how the computed $\alpha$ is affected by numerical resolution.

@@

While detailed three-dimensional hydrodynamic modeling of protostellar disks can provide considerable insight into the physical and thermal states of these disks, computing resources limit such studies to snapshots spanning timescales much shorter than a disk's evolutionary lifetime.  In contrast, approaches that replace detailed physics with statistical parameters that encapsulate this physics allow for calculations covering much longer timeframes.  Both approaches can play important roles in broadening our understanding of these complex nonlinear systems, provided a connection between the detailed physics and statistical parameters can be demonstrated and understood. In what follows, we examine the applicability of an $\alpha$-disk formulation \citep{shakura1973} to  the
evolution of protostellar disks subject to gravitational instabilities.

Gravitational instabilities (GIs) play an important, {\bf SM: I'm not sure this is true in general. Perhaps reword to ``GIs can play an important role...and may dominate the structure for certain phases of evolution''}and at times dominant, role in driving the structural and thermal
evolution of  protostellar disks \citep[for a review, see][]{durisen2005}.
%In particular, torques arising from GIs provide a very effective means of redistributing matter and 
%angular momentum within these disks 
%\citep{Lod04, Lod05, Vor05, Vor06, Bol09}.  
Thermal processes play the primary role in regulating the amplitude and outcome of these instabilities \citep{pickett1998,pickett2000,mejia2005,nelson1998,nelson2000a}. A a disk's susceptibility to GIs is parameterized by the Toomre $Q$-parameter \citep{toomre1981};  $Q = c_s \kappa / \pi G \Sigma$, where $c_s$ is the sound speed, $\kappa$ is the epicyclic frequency ($\sim$ the rotation frequency 
$\Omega$ in a nearly Keplerian disk), and $\Sigma$ is the disk surface mass density. For $Q \al$  1.5 to 1.7,  small density perturbations in a disk grow exponentially on a time scale comparable to the rotation period. These perturbations manifest themselves as multi-arm spirals with a predominantly trailing pattern that transports angular momentum outward by gravitational torques \citep[e.g.]{larson1984,boss1984,durisen1986,papaloizou1991, laughlin1994,nelson1998,pickett1998}. The long range nature of these torques cause nonlinear
spiral structures to develop over a broad range of radii \citep{laughlin1998,nelson1998,nelson2000a,pickett1998, pickett2000,pickett2003}. {\bf SM: not sure of the meaning here. A low Q is what causes GIs to develop over a broad range of radii. Perhaps you mean the long range nature can cause coherent spiral structures to form with large radial extent?}

Numerous studies have shown that thermal physics control GI amplitudes by a {\it balance} of heating and cooling \citep[for example][]{tomley1991,tomley1994,pickett1998,pickett2000, pickett2003,gammie2001,boss2002,rice2003b,mejia2005,boley2006,boley2007, stamatellos2008,cossins2009}.
Heating by GIs tends to increase $c_s$, thus increasing $Q$. If disk heating persists, the instability will be suppressed once $Q$ becomes sufficiently large.  However, heating and cooling can reach a balance at nonlinear wave amplitude. GI-activity can sustain this balance  at a relatively constant,
but unstable, value of $Q$ \citep{paczynski1978, lin1981,goldreich1965},and 
the disk can exist in a state of quasi-equilibrium. In this state cooling is 
balanced by heating arising from the inward flow of matter and shocks induced by the GIs
\citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime \citep[e.g.][]{nelson1998, pickett1998} {\bf additional refs?}.  Even if the initial growth is dominated by one mode, numerous modes usually appear in the nonlinear regime. As a consequence, if a disk achieves a quasi-steady balance between heating and cooling, then there tends to be power at all $m$-values resolved by a given numerical method \citep{mejia2005,boley2006}.  
In this  {\it asymptotic state}, GI-active disks develop overlapping density structures of different strengths, geometries and coherence, producing strongly nonlinear turbulence that pervades the entire disk \citep[e.g.][]{pickett2003, mejia2005,boley2006}.
\citet{gammie2001} refers to this state as {\it gravitoturbulence}. Gravitoturbulence provides a possible bridge between the detailed physics of GIs and the viscous transport parameter $\alpha$.

\citet{shakura1973} first parameterized turbulent transport in a steady-state disk with the  parameter $\alpha$, where the source of the turbulence is unspecified. Although their original work was directed towards X-ray bright
accretion disks around black holes, this heuristic approach has proved fruitful in many astrophysical situations by allowing for analytic solutions and relatively easy numerical modeling over time-spans inaccessible to detailed modeling. In the $\alpha$ disk formalism, when the accretion rate $\dot{M}$ is constant in both time and radius, $\dot{M} = 3\pi\alpha c_s \Sigma H$ where $c_s$ is the sound speed, $\Sigma$ the surface density, and $H$ the disk scale height. The $\alpha$ formalism makes no assumption about the nature of the angular momentum transport, only that {\it heating and cooling are dominated by local processes}{\bf SM: This is not my understanding of the $\alpha$ formalism, I don't think it says anything about cooling. It also means that ang mom transport happens locally and that this causes local heating}; the validity of an $\alpha$-disk picture requires self-consistency of local energy dissipation.

Dissipative locality {\bf SM: need to explain what is meant by this more precisely} in GI-active disks has not been rigorously tested and remains an open question. 
Several studies support the idea that GI transport in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006}, and disks observed in the embedded phase can be quite massive relative to their central stars \citep[e.g.][]{osorio2003}. 
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in balance it may be valid to represent GI-induced transport by an {\it effective viscosity}, a local process \citep[e.g.][]{gammie2001, rice2005, lodato2004}. Indeed, \citet{gammie2001} derived an effective alpha for the case of a thin 2D gravitationally unstable disk where the heating due to local turbulent dissipation caused by GIs was balanced by the local cooling rate.  Modifications to this derivation have since been proposed by {\bf (Tom: include other derivations and refs)}.   Recent simulations \citep[e.g.]{cossins2009, vorobyov2010} suggest that, at least in the case of imposed cooling, $\alpha_{\rm{eff}}$ roughly converges to the value predicted by \citeauthor{gammie2001}.

{\bf (Tom: rework and finish this paragraph, or toss it)}
Three possibilities exist: 1) GI-transport is predominately a long-range effect and thus not 
amenable to an $\alpha$-disk representation, 2) GI-transport can be well represented as
a short-range effect and thus is directly amenable to an $\alpha$-disk representation,
If long-range energy transport and/or angular momentum transport are negligible,  or if
the long-range transport could be shown to represent a known fraction of the 
***. 
Identifying the principal angular momentum transport mechanism
and how it behaves is critical to understanding whether disks can be described by an alpha disk model.

{\bf (Tom: add text concerning need for convergence approach)}
Here we report the results of a convergence study designed to determine whether GI-active disks, chaotic by nature, achieve statistical equilibrium; if so, to what level do these disks approximate the gravitoturbulence of \citet{gammie2001} and with what distribution of $\alpha_{eff}$; and to characterize how the GI cascades from low-order to high-order modes. To this end, we follow the evolution of a GI-active disk, subject to constant cooling, in its asymptotic state, where cooling and heating are essentially balanced.  Simulations are performed using four different resolutions in the azimuthal direction.

{\bf (Tom: add paragraph providing outline of paper)}

%Nonlinear evolutions of disks undergoing GIs, even local thin-disk treatments, 
%\citep[e.g.,][]{Joh03} 
%are sufficiently complex that they must be done numerically,  
%typically using a finite element approach, where
%a set of coupled equations is solved simultaneously on a computational grid.  
%Computational constraints prohibit a full examination of GI unstable disks over 
%the life of the disk.
%Indeed, existing computational resources limit 
%detailed three-dimensional 
%simulations of disk evolution to tens of
%outer disk rotations.  
%We cannot yet follow the entire dynamical evolution of a 
%protoplanetary disk nor
%model all the radiative and particle processes of interest at once in full 3D. However, 
%by following GI simulations to their asymptotic states, when these exist, we
%can extrapolate the consequences over much longer times (e.g., Boley et al 2006).

%For 
%grids of a fixed {\it physical} size,
%grids with more elements provide finer meshes, which typically produce a more accurate solution. 
%However, this increased accuracy if offset by temporal sacrifices.
%The number of equations to be solved at each timestep 
%increases as the grid size is increased.  At the same time, the Courant condition 
%forces these timesteps to be shorter in duration.  A mesh convergence study 
%provides a common approach to balancing accuracy and computing resources.
%In the following sections we follow the evolution of a disk subject to GUs
%using four different
%resolutions in the azimuthal direction.  The number of azimuthal grid points used in simulations
%may be particularly important, because higher resolution 
%in azimuth allows GI power to spread to higher-order modes 
%that behave more locally. We test this by examining the power in various 
%modes to see how they are affected by the choice of grid.

@@

Figure \ref{Final_Q} shows $Q$, as a function of radius, at $t \sim$ 18 ORP, while table \ref{tbl:ams} lists $Q_{avg}$, the value of $Q$ of for the asymptotic disk, time-averaged of 12 -- 19 ORP and spatially averaged over 10 -- 40 AU, as a function of radius for each resolution,

@@

Figure \ref{fig:DensityPlots}, shows midplane and radial densities  at $t \sim 18$ ORPs for the $l_{max} =$ 64, 128, 256, and 512 simulations. The lower resolution (smaller $l_{max}$) simulations display longer wavelength spiral structures, with larger amplitudes and more coherence than those seen in higher resolution simulations. Higher angular resolution allows GI power to spread to higher-order azimuthal modes, as demonstrated by a Fourier decomposition of the density. The strengths of these modes are given by their global Fourier amplitudes,

@@

Here  $\rho_o$ is the axisymmetric (zero-order)  component of the density. In what follows, $A_m$ is averaged over time to suppress fluctuations on the dynamic time scale due to the chaotic nature of GIs in the asymptotic  phase.
In particular, $\langle A_{tot} \rangle$ will represent the power over the radial range 10 -- 40 AU, integrated from $ m = 2$ to $l_{max}/2$ and averaged over the time range 12 -- 19 ORPs. The Fourier amplitudes cannot be accurately measured at resolutions smaller than $l_{max}/2$ \citep{shannon1984}, therefore $A_m$ is limited to modes $m \le l_{max}/2$.  
Because the central star is artificially fixed to the grid center, the $m = 1$ mode cannot be be realistically treated; it is excluded from $\langle A_{tot} \rangle$.

@@

Previous studies have shown that disks in the asymptotic state have power at all resolvable $m$-values  \citep[e.g.][]{mejia2005,boley2006}.  Table \ref{tbl:ams} lists $\langle A_{tot} \rangle$ along with $\langle A_{2-7} \rangle  /  \langle A_{tot} \rangle$,  the fraction of the total power falling into low-order ($m = 2$ -- 7) modes, for each resolution. While the total power is approximately independent of resolution, $\langle A_{tot} \rangle = 1.90 \pm 0.07$, 
the distribution of this power shows marked differences. In particular, the fractional power in lower order modes decreases monotonically with increasing resolution.  Over three quarters of $\langle A_{tot} \rangle$ resides
in modes $m =$ 2 -- 7 for the $l_{max} = 64$ simulation, while approximately half resides in these modes when $l_{max} =$ 512.

The shift of power from lower to higher order modes as the resolution of the computational grid is increased arises from the fact that, while $\langle A_{tot} \rangle$ is approximately independent of $l_{max}$, {\bf SM: Need some explanation as to why this is the case.} the number of degrees of freedom available for this power, $l_{max}/2$, increases linearly with the resolution. Thus power is naturally spread from lower order to higher order modes as the available $m$-values increases. Indeed, even if the disk is dominated by one or two low-order modes at the time of outburst, this power cascades to higher-order modes \citep[for example][]{laughlin1997, laughlin1998,laughlin1996}.

Given the resolution dependence of how power is distributed,  we must examine what resolution is sufficient to properly capture the evolution of these disks. Table \ref{tbl:ams} shows that while the fractional power in lower order modes decreases with each doubling of $l_{max}$, the amount of change in going from $l_{max} = 256$ to 512 
is significantly less than the change going from $l_{max} = 64$ to 128, with a strong suggestion that this diminution will not proceed much further with higher azimuthal resolution. Indeed, a  Richardson extrapolation {\bf (add reference) SM: suggest Press et. al} \citep{press1992} of the tabulated values of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ suggests a limiting value of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle \sim$ 0.43 for increasingly higher resolutions. This is displayed graphically in figure 4, where the fractional power in lower order modes is plotted against the  normalized grid 
spacing for each value of $l_{max}$. The grid spacing  is normalized such that spacing of the $l_{max}$ = 512 grid
equals unity.  Convergence occurs as the normalized spacing goes to zero. The Richardson extrapolation value  of $\langle A_{2-7}\rangle  / \langle A_{tot}\rangle  = 0.43$ is plotted at a normalized grid spacing  equal to zero. More than half of the modal power in a converged disk resides in higher-order modes, modes that help define a state of gravitoturbulence. {\bf SM: I'm not sure this is accurate, gravitoturbulence isn't really turbulence (at least as it's normally defined) and the high order modes aren't really turbulence either} The fact that the $l_{max} =$ 512 simulation has $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ near the convergence value lends some confidence in the use of the $l_{max} =$ 512 model for the studie of effective $\alpha$.

@@

Coherent density structures in a disk produce gravitational torques. Low-order modes in $\Delta\rho / \rho$  have longer wavelengths and, if coherent in radius, have longer lever arms to produce torques. On the other hand, high-order modes have relatively short wavelengths which, particularly if they lack coherence, produce more localized effects. The high-order modes tend to produce pockets of local GI activity, which can cancel each other
over the whole disk, {\bf SM: I think there's a sentence missing here explaining that higher resolution leads to more of the amplitude being concentrated in high order modes, while the total amplitude stays the same.} leading to the expectation that  gravitational torques in the disk will decrease as $l_{max}$ is increased. This can be tested by examining the torque contributions from  various Fourier modes.

@@

When considering the effect of azimuthal resolution on the asymptotic behavior
of a GI unstable disk I found that without sufficient azimuthal resolution of at least
$l_{max} =$ 512 gravitational torques and effective gravitational $\alpha$
will be underestimated
for disk with a constant global cooling rate. I also found that, when properly resolved,
more non-axisymmetric power propagates to higher order modes, and the amplitude
of low-order modes decreases. This results in a decrease of the gravitational torque
and effective gravitational $\alpha$. For $l_{max} =$ 512 the effective gravitational 
$\alpha$ measured
is in very good agreement with the prediction of Gammie (2001). Gammie's 
prediction follows from an argument of balance of GI heating and cooling, resulting in
``gravitoturbulence''. However, his detailed simulations found that the torques and
heating in the disk were due in equal parts to gravitational stresses and Reynolds
stresses, this is in contrast to my measurements which only consider gravitational
stresses.

\acknowledgements

@@

\begin{figure}\label{Final_Q}
\plotone{figures/qcompareres.eps}
\caption{Toomre Q distribution versus radius for each of the four resolutions at t $\sim$ 18 ORPs}
\end{figure}

\begin{figure}\label{fig:DensityPlots}
%\plotone{figures/Figure_1.eps}
\caption
{
Midplane disk densities at $t \sim 18$ ORPs for simulations with 64, 128, 256 and 512 azimuthal grid elements.
The simulations differ only in the number of  azimuthal zones used in the calculations.  Densities out of the plane, along an azimuthal cut through the disk, are displayed below the face-on views.  The color scale is  logarithmic, and axis units are AU. 
}
\end{figure}
\newpage

>>>>

\title{Convergence Studies on Mass Transport in Disks with Gravitational Instabilities.
 \\
I. The Constant Cooling Case}
\author{Scott Michael}
\affil{Astronomy Department, Indiana University, Bloomington, IN 47405}
\email{scamicha@astro.indiana.edu}

@@

We conducted a 3D hydrodynamic convergence study of a protostellar disk
subject to a constant global cooling time 
and to gravitational instabilities (GIs). 
The disk was followed during its asymptotic state, where
heating and cooling are in balance.
Four simulations were conducted, identical except for the number of azimuthal grid 
points, $l_{max} = $ 64, 128, 256 and 512, used in the calculations. 
The strengths of the azimuthal modes, gravitational torques, and the effective \citet{shakura1973} $\alpha$ 
arising from gravitational stresses were determined for each resolution.  These
were used to examine the resolution-dependence of each quantity
and to evaluate both the converged modal power distribution and the effective value of $\alpha$
as $l_{max} \rightarrow \infty$.   Power was found in all modes and 
the power summed over all modes was essentially independent of resolution.
The modal distribution of power, however, was significantly affected by resolution.  
The size of the azimuthal grid, $l_{max}$,
limits the number of modes to $l_{max}/2$.  Higher resolution simulations therefore
have more modes available to them.  Power is spread across all available modes.  As a consequence,
higher resolution simulations have a larger fraction of their total power in high-order modes.
High-order modes
manifest themselves more locally than low-order modes, therefore 
greater resolution leads to smaller gravitational stresses.
As a result, higher resolution simulations experience 
weaker gravitational torques than lower resolution simulations.   The effective  $\alpha$
also depends upon the magnitude of the stresses.  Thus
$\alpha_{eff}$ is also a function of grid resolution, decreasing as  $l_{max}$ increases.
The converged $\alpha_{eff}$ is consistent with
predictions from an analytic local theory for thin disks by \citet{gammie2001}, 
but only over many dynamic times when averaged over a substantial volume of the disk.
Results obtained in the $l_{max} = 512$ simulations differ with converged values
by less than 20\%.

@@

Detailed three-dimensional hydrodynamic modeling of protostellar disks can provide considerable insight into the physical and thermal states of these disks.  However,  computing resources limit these studies to snapshots spanning timescales much shorter than a disk's evolutionary lifetime. The level of spatial resolution (and hence temporal resolution, due to the Courant condition) used in calculations can play a deterministic role in interpretations of models; while lower resolutions allow more extensive spatial and temporal studies, this benefit can come at the sacrifice of important physics.  Convergence studies provide a means 
of deriving important outcomes without the expense of very high resolution simulations.

In contrast to sophisticated calculations that include detailed physics, 
approaches exist that replace detailed physics with statistical parameters that encapsulate this physics.  This
allows for calculations covering much longer timeframes.  Both detailed and 
parametric
approaches can play important roles in broadening our understanding of complex nonlinear systems, provided a connection between the detailed physics and statistical parameters can be demonstrated and understood. In what follows, we perform convergence 
tests on hydrodynamical disk simulations to examine the applicability of an $\alpha$-disk 
formulation \citep{shakura1973} to  the
evolution of protostellar disks subject to gravitational instabilities.

%Here $\alpha$ is a dimensionless parameter that
%provides a measure of viscosity, without providing a detailed specification of the viscous mechanism.

Gravitational instabilities (GIs) can play an important, and at times dominant, role in driving the structural 
evolution of  protostellar disks \citep[for reviews, see][]{durisen2005, durisen2010}. {\bf (We need the Durisen 2010 (2011?)
review article reference)}
Thermal processes play the primary role in regulating the amplitude and outcome of these instabilities \citep{pickett1998,pickett2000,mejia2005,nelson1998,nelson2000a}. A disk's susceptibility to GIs can be parameterized by the Toomre $Q$-parameter \citep{toomre1981};  $Q = c_s \kappa / \pi G \Sigma$, where $c_s$ is the sound speed, $\kappa$ is the epicyclic frequency ($\sim$ the rotation frequency 
$\Omega$ in a nearly Keplerian disk), and $\Sigma$ is the disk surface mass density. For $Q \al$  1.5 to 1.7,  small density perturbations in a disk grow exponentially on a time scale comparable to the rotation period. These perturbations manifest themselves as multi-arm spirals with a predominantly trailing pattern that transports angular momentum outward by gravitational torques \citep{larson1984,boss1984,durisen1986,papaloizou1991, laughlin1994,nelson1998,pickett1998}.

Numerous studies have shown that thermal physics control GI amplitudes by a {\it balance} of heating and cooling \citep[e.g.,][]{tomley1991,tomley1994,pickett1998,pickett2000, pickett2003,gammie2001,boss2002,rice2003b,mejia2005,boley2006,boley2007, stamatellos2008,cossins2009}.
Heating by GIs tends to increase $c_s$, thus increasing $Q$. If disk heating persists, the instability will be suppressed once $Q$ becomes sufficiently large.  However, heating and cooling can reach a balance at nonlinear wave amplitude. GI-activity can sustain this balance  at a relatively constant,
but unstable, value of $Q$ \citep{paczynski1978, lin1981,goldreich1965}, and 
the disk can exist in a state of quasi-equilibrium. In this state, cooling is 
balanced by heating arising from the inward flow of matter and shocks induced by the GIs
\citep{gammie2001, lodato2004, rice2005, boley2006, cossins2009, vorobyov2010}.

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime \citep{nelson1998, pickett1998, lodato2004,boley2006,cossins2009}.  Even if the initial growth is dominated by one mode, numerous modes usually appear in the nonlinear regime. As a consequence, if a disk achieves a quasi-steady balance between heating and cooling, referred to as the {\it asymptotic state},
then power will exist at all $m$-values resolved by a given numerical method \citep{mejia2005,boley2006}.  
In the asymptotic state, GI-active disks develop overlapping density structures of different strengths, geometries and coherence, producing strongly nonlinear turbulence that pervades the entire disk \citep[e.g.][]{pickett2003, mejia2005,boley2006}.
\citet{gammie2001} refers to this state as {\it gravitoturbulence}. Gravitoturbulence provides a possible bridge between the detailed physics of GIs and the viscous transport parameter $\alpha$.  Here convergence studies are important because 
higher order modes dissipate energy on shorter wavelengths than lower order modes.  The mix of low and high order modes available to a disk is set by the resolution of the simulation.

\citet{shakura1973} proposed turbulence as the primary source of effective viscosity in gas disks
and parameterized turbulent transport in a steady-state disk with the  parameter $\alpha$.  The source of the 
turbulence in their $\alpha$-disk formulation is unspecified. Although their original work was directed towards X-ray bright
accretion disks around black holes, this heuristic approach has proven fruitful in many astrophysical situations by allowing for analytic solutions and relatively easy numerical modeling over time-spans inaccessible to detailed modeling. In the 
$\alpha$-disk formalism, when the accretion rate $\dot{M}$ is constant in both time and radius, $\dot{M} = 3\pi\alpha c_s \Sigma H$ where $\Sigma$ is the surface density and $H$ represents
the disk scale height. The $\alpha$ formalism makes no assumption about the nature of the angular momentum transport, only that {\it angular momentum transport
and heating are dominated by local processes}. Lin and Pringle (1987, 1990) were among the first to suggest
that GI-induced transport could be described within a viscous
framework.  However, the  validity of an $\alpha$-disk picture requires self-consistency of local energy dissipation. 
The locality of dissipation in GI-active disks has not been rigorously tested and remains an open question.

Several studies support the idea that GI transport in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006}.  
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in balance, i.e., disks subject to gravitoturbulence,
 it may be valid to represent GI-induced transport by an {\it effective viscosity}, a local process \citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.  In particular, 
an $\alpha$-viscosity may be consistent
with transport in GI acitve disks for disk to star mass ratio $M_d/M_* \le 0.2$ -- 0.3; GI-active disks
in systems with larger $M_d/M_*$ are dominated by lower-order modes that act in a more global sense and thus cannot be well represented by an $\alpha$-approach
\citep{lodato2004, cossins2009, vorobyov2010}.

\citet{gammie2001} derived an effective alpha for the case of a thin 2D gravitationally unstable disk where the heating due to local turbulent dissipation caused by GIs was balanced by the local cooling rate.  
Assuming a razor thin gravitationally unstable disk where the heating due to local turbulent dissipation caused by
GIs was balanced by a local cooling rate of $t_{cool}$.
Gammie found
$\alpha_{eff} = [\gamma(\gamma - 1) \slantfrac{9}{4}\Omega t_{cool}]^{-1}$,
where $\gamma$ is the two dimensional adiabatic index and $t_{cool}$ is a cooling parameter. 
The two dimensional index relates
to the three dimensional index, $\Gamma$, by 
$\gamma = (3\Gamma - 1) / (\Gamma + 1)$
for a non-self-gravitating
disk and
$\gamma = 3 - 2/\Gamma $
for a strongly self gravitating disk.   \cite{lodato2004} derived essentially the same result for their 3D study.
Motivated by
the \cite{laughlin1996}, \cite{gammie2001}, and \cite{lodato2004, lodato2005} results 
that demonstrated differences in how low and high order modes manifest themselves, Kratter et al. (2008) 
proposed $\alpha_{eff} = (\alpha_{short}^2 + \alpha_{long}^2)^{1/2}$, where $\alpha_{short}$ and $\alpha_{long}$
are functions of $Q$ and $M_d/M_*$ representing 
the different wavelength regimes expected to dominate for different values of the mass ratio and cooling rate. 
{\bf (Scott: need to add kratter2008 to the bib file.  The reference is Kratter, K.M., Matzner, C. D., Krumholz, M.R., 
2008. ApJ 681, 375)}
Fully 3D simulations \citep[e.g.][]{lodato2004, cossins2009} suggest that, at least in the case of imposed cooling, $\alpha_{\rm{eff}}$ roughly converges to the value predicted by \citeauthor{gammie2001}.

We address these, and similar questions, using a grid-based, finite-difference 3D hydrodynamics code
with self gravity.
For computational grids of a fixed {\it physical} size,
grids with more elements provide finer meshes, and thus a more accurate solution. 
However, increased accuracy is offset by temporal sacrifices.
The number of equations solved at each timestep 
increases as the grid size is increased.  At the same time, the Courant condition 
forces shorter timesteps.  A mesh convergence study 
provides a common approach to balancing accuracy and computing resources.
In the following sections we follow the evolution of a disk subject to GIs
using four different
resolutions in the azimuthal direction.  The number of azimuthal grid points used in simulations
may be particularly important, because higher azimuthal resolution 
allows GI power to spread to higher-order modes 
that behave more locally. We test this by examining the power in various 
modes to see how they are affected by the choice of grid.
In the process, 
we report the results of a convergence study designed to determine whether GI-active disks, chaotic by nature, achieve statistical equilibrium; if so, to what level do these disks approximate the gravitoturbulence of \citet{gammie2001} and with what distribution of $\alpha_{eff}$; and to characterize how the GI cascades from low-order to high-order modes. To this end, we follow the evolution of a GI-active disk, subject to constant cooling, in its asymptotic state, where cooling and heating are essentially balanced.  Simulations are performed using four different resolutions in the azimuthal direction.

The balance of this paper is organized as follows.  Section 2 provides the details of the numerical approach and defines the models.  An analysis of the modal structure of the simulations 
is found in \S3.1, the gravitational torques arising from these modes follows in \S3.2.  Effective converged alphas are 
discussed in \S3.3.  Section 3.4 deals with the degree to which selected grid sizes lead to results consistent with
convergence.  Fragmentation is discussed in \S3.5.  A summary of results is found in \S4.

%We cannot yet follow the entire dynamical evolution of a 
%protoplanetary disk nor
%model all the radiative and particle processes of interest at once in full 3D. However, 
%by following GI simulations to their asymptotic states, when these exist, we
%can extrapolate the consequences over much longer times (e.g., Boley et al 2006).

@@

Figure 
%\ref{qcompareres}
\ref{fig:Final_Q} 
shows $Q$, as a function of radius, at $t \sim$ 18 ORP.  Table \ref{tbl:ams} lists $Q_{avg}$, the value of $Q$ of for the asymptotic disk, time-averaged of 12 -- 19 ORP and spatially averaged over 10 -- 40 AU, as a function of radius for each resolution.  From $\sim$ 12 -- 50 AU, $Q \sim 1.0$ -- 1.4 and thus the disk is subject to the growth of nonaxisymmetric structures over this full range.  Interior to $\sim$ 27 AU, the value of $Q$ shows no dependence on
the azimuthal resolution.  From 27 -- 50 AU, lower resolution simulations display lower values of $Q$.

@@

Figure \ref{fig:DensityPlots}, shows midplane and radial densities  at $t \sim 18$ ORPs for the $l_{max} =$ 64, 128, 256, and 512 simulations. Two features are readily apparent; disks run at higher resolution are thicker than lower resolution disks {\bf (anyone want to comment on this?)} and the character of density perturbations differ with resolution.
The lower resolution (smaller $l_{max}$) simulations display longer wavelength spiral structures, with larger amplitudes and more coherence than those seen in higher resolution simulations. Higher angular resolution allows GI power to spread to higher-order azimuthal modes, as demonstrated by a Fourier decomposition of the density. The strengths of these modes are given by their global Fourier amplitudes,

@@

Here  $\rho_o$ is the axisymmetric component of the density. In what follows, $A_m$ is averaged over time to suppress fluctuations on the dynamic time scale due to the chaotic nature of GIs in the asymptotic  phase.
In particular, $\langle A_{tot} \rangle$ will represent the power over the radial range 10 -- 40 AU, integrated from $ m = 2$ to $l_{max}/2$ and averaged over the time range 12 -- 19 ORPs. The Fourier amplitudes cannot be accurately measured at resolutions smaller than $l_{max}/2$ \citep{shannon1984}, therefore $A_m$ is limited to modes $m \le l_{max}/2$.  
Because the central star is artificially fixed to the grid center, the dynamics of the $m=1$ mode may not
be accurately treated. For this reason, it is excluded from $\langle A_{tot} \rangle$.

@@

Previous studies have shown that disks in the asymptotic state have power at all resolvable $m$-values  \citep{lodato2004,mejia2005,boley2006,cossins2009}.  Table \ref{tbl:ams} lists $\langle A_{tot} \rangle$ along with $\langle A_{2-7} \rangle  /  \langle A_{tot} \rangle$,  the fraction of the total power falling into low-order ($m = 2$ -- 7) modes, for each resolution. While the total power is approximately independent of resolution, $\langle A_{tot} \rangle = 1.90 \pm 0.07$, 
the distribution of this power shows marked differences. In particular, the fractional power in lower order modes decreases monotonically with increasing resolution.  Over three quarters of $\langle A_{tot} \rangle$ resides
in modes $m =$ 2 -- 7 for the $l_{max} = 64$ simulation, while approximately half resides in these modes when $l_{max} =$ 512.

The shift of power from lower to higher order modes as the resolution of the computational grid is increased arises from the fact that, while $\langle A_{tot} \rangle$ is approximately independent of $l_{max}$,  the number of degrees of freedom available for this power, $l_{max}/2$, increases linearly with the resolution. Thus power is naturally spread from lower order to higher order modes as the available $m$-values increases. Indeed, even if the disk is dominated by one or two low-order modes at the time of outburst, this power cascades to higher-order modes \citep[for example][]{laughlin1997, laughlin1998,laughlin1996}.

Given the resolution dependence of how power is distributed,  we must examine what resolution is sufficient to properly capture the evolution of these disks. Table \ref{tbl:ams} shows that while the fractional power in lower order modes decreases with each doubling of $l_{max}$, the amount of change in going from $l_{max} = 256$ to 512 
is significantly less than the change going from $l_{max} = 64$ to 128, with a strong suggestion that this diminution will not proceed much further with higher azimuthal resolution. Indeed, a  Richardson extrapolation \citep{press1992} of the tabulated values of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ suggests a limiting value of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle \sim$ 0.42 for increasingly higher resolutions. This is displayed graphically in Figure \ref{fig:Power_Extrap}, where the fractional power in lower order modes is plotted against the  normalized grid 
spacing for each value of $l_{max}$. The grid spacing  is normalized such that spacing of the $l_{max}$ = 512 grid
equals unity.  Convergence occurs as the normalized spacing goes to zero. The Richardson extrapolation value  of $\langle A_{2-7}\rangle  / \langle A_{tot}\rangle  = 0.42$ is plotted at a normalized grid spacing  equal to zero. The majority of the modal power in a converged disk resides in higher-order modes. 
%These modes dissipate on shorter wavelengths than low-order modes and hence act more locally.

@@

Low-order spiral modes in $\Delta\rho / \rho$  have longer wavelengths that, if coherent in radius, have longer lever 
arms to produce torques, while high-order modes have relatively short wavelengths which, particularly if they 
lack coherence, produce more localized effects. High-order modes tend to produce pockets of local GI activity, 
which can cancel each other over the whole disk.  The finding that the fractional power in high-order modes increases
with resolution, while the total power stays the same, leads to
 the expectation that  gravitational torques in the disk will decrease as $l_{max}$ is increased. This can be tested by examining the torque contributions from  various Fourier modes.

@@

\noindent
{\bf (All: My style of writing papers is to write the summary after everything is completed.  Therefore, I have left this
section blank.)}

\acknowledgements
Please add any acknowledgements here.

@@

\begin{figure}
%\plotone{figures/qcompareres.eps}
\plotone{figures/qcompareres.pdf}
\caption{Toomre Q distribution versus radius for each of the four resolutions at t $\sim$ 18 ORPs}
\label{fig:Final_Q}
\end{figure}

\begin{figure}
%\plotone{figures/Figure_1.eps}
\plotone{figures/Figure_1.pdf}
\caption
{
Midplane disk densities at $t \sim 18$ ORPs for simulations with 64, 128, 256 and 512 azimuthal grid elements.
The simulations differ only in the number of  azimuthal zones used in the calculations.  Densities out of the plane, along an azimuthal cut through the disk, are displayed below the face-on views.  The color scale is  logarithmic, and axis units are AU. 
}
\label{fig:DensityPlots}
\end{figure}
\newpage
